### 2024-06-05 journal/algorithm.tex
some edits from final proofreading

%\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ performs the following
three steps.
$P$ first observes
the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
\rev{Using terminology from \cite{DBLP:conf/fsttcs/JayantiP05}, we call the procedure that does} this three-step sequence a
\op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on~$\var{v}$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $\var{v}$'s children to $\var{v}$,
we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
then $P$'s operation has been propagated to $\var{v}$ by some other process, so $P$ can continue 
onwards towards the~root.

@@

\begin{figure*}[t]
\input{tournamentTree.pdf_t}
\caption{An example \ordering\ tree with four processes. 
E(x) denotes an Enqueue(x) operation and D\sub{1} to D\sub{6} denote \dequeue\ operations.
We show explicitly the \enqueue\ sequence and \dequeue\ sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's D\sub{6} is still propagating towards the root.
The linearization order for this tree is
E(a) E(e) D\sub{2} $\mid$ E(b) D\sub{4} D\sub{5} $\mid$ E(d) E(f) E(h) D\sub{1} $\mid$ E(c) D\sub{3} $\mid$ E(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
\medskip
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{orderingtree}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as in Figure \ref{orderingtree}.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure*}

@@

Performing a double \op{Refresh} at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation completes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in the root's \fld{blocks} array.
Thus, we linearize operations according to the block they belong to in the root's array.
We can choose how to order operations within the same block, since they must be concurrent.

@@

Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child.  
Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].\hspace*{3mm}
\end{eqnarray}
The \eleft\ and \eright\ fields allow us to navigate to a block's direct subblocks.
Blocks also store some prefix sums:
%the block in 
$\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total numbers of \enqueues\ and \dequeues\ in $\var{v}.\fld{blocks}[1..\var{b}]$.
We use these to search for a particular operation.
% pinpoint the  location of an operation among the subblocks of a given block.
For example, consider finding the $r$th \enqueue\ $E_r$ in the linearization.
A binary search for $r$ on the \fld{sum\sub{enq}} fields of the root's blocks 
finds the block  containing $E_r$.
If we know a block $B$ in a \node\ $\var{v}$ contains $E_r$,
we can use the \fld{sum\sub{enq}} field again to determine which child of $\var{v}$ contains $E_r$
and then do a binary search
of the \fld{sum\sub{enq}} fields of the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find the leaf block that  stores 
$E_r$ explicitly.

\rev{Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.}
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf takes $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
%All information needed for this search process can be derived from the 
%\eleft, \eright\ and \fld{sum\sub{enq}} fields.

>>>>

%\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ performs the following
three steps.
$P$ first observes
the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
\rev{Using terminology from \cite{DBLP:conf/fsttcs/JayantiP05}, we call the procedure that does} this three-step sequence a
\op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on~$\var{v}$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $\var{v}$'s children to $\var{v}$,
we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
then $P$'s operation has been propagated to $\var{v}$ by some other process.
\rev{Thus, after performing a double \op{Refresh},} $P$ can continue 
onwards towards the~root.

@@

\begin{figure*}[t]
\input{tournamentTree.pdf_t}
\caption{An example \ordering\ tree with four processes. 
E(x) denotes an \enqueue(x) operation and D\sub{1} to D\sub{6} denote \dequeue\ operations.
We show explicitly the \enqueue\ sequence and \dequeue\ sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's D\sub{6} is still propagating towards the root.
The linearization order for this tree is
E(a) E(e) D\sub{2} $\mid$ E(b) D\sub{4} D\sub{5} $\mid$ E(d) E(f) E(h) D\sub{1} $\mid$ E(c) D\sub{3} $\mid$ E(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
\medskip
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{orderingtree}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as in Figure \ref{orderingtree}.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure*}

@@

Performing a double \op{Refresh} at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation completes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in the root's \fld{blocks} array.
Thus, we linearize operations according to the block they belong to in the root's array.
We can freely choose how to order operations within the same block, since they must be concurrent.

@@

Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child, \rev{respectively}.  
Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}}+1..\nonumber\\
		\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].\hspace*{3mm}
\end{eqnarray}
The \eleft\ and \eright\ fields allow us to navigate to a block's direct subblocks.
Blocks also store some prefix sums:
%the block in 
$\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total numbers of \enqueues\ and \dequeues, \rev{respectively}, in $\var{v}.\fld{blocks}[1..\var{b}]$.
We use these to search for a particular operation.
% pinpoint the  location of an operation among the subblocks of a given block.
For example, consider finding the $r$th \enqueue\ $E_r$ in the linearization.
A binary search for $r$ on the \fld{sum\sub{enq}} fields of the root's blocks 
finds the block  containing $E_r$.
If we know a block $B$ in a \node\ $\var{v}$ contains $E_r$,
we can use the \fld{sum\sub{enq}} field again to determine which child of $\var{v}$ contains $E_r$
and then do a binary search
of the \fld{sum\sub{enq}} fields of the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find the leaf block that  stores 
$E_r$ explicitly.

\rev{Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.}
We shall show \rev{in Section \ref{sec-analysis}} that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf takes $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
%All information needed for this search process can be derived from the 
%\eleft, \eright\ and \fld{sum\sub{enq}} fields.

### 2024-06-05 journal/preliminaries.tex
some edits from final proofreading

\rev{
We use a  standard model of an asynchronous, shared-memory system.
A collection of $p$ processes can communicate by accessing
shared memory locations using read, write and compare-and-swap (CAS) instructions.
A \op{CAS}(\var{X}, \var{old}, \var{new}) instruction updates a variable \var{X} to the value \var{new} and returns \tr\ if \var{X}'s value is \var{old}; otherwise the \op{CAS} leaves \var{X} unchanged and returns \fa.
We also assume processes can allocate a block of shared memory to represent an object.

We are concerned with implementing a first-in, first-out queue in such a system,
which requires an algorithm that a process can execute to perform
an enqueue or dequeue operation.
We treat individual memory instructions as atomic, so an execution
of such an implementation can be thought of as a sequence of steps by processes
where in each step a process performs at most one shared-memory instruction 
(and receives the response of that instruction).
The subsequence of steps performed by a process should follow that process's
algorithm for performing a sequence of enqueue and dequeue operations.
The order in which the steps of processes are interleaved is arbitrary, and the implementation must be correct for all
possible interleavings.

@@

The step complexity of an operation is the number of steps it performs.
The amortized step complexity is the total number of steps by all processes
divided by the number of operations they perform.
More formally, to show that the amortized step complexity is $O(T_e)$ per enqueue
and $O(T_d)$ per dequeue, we must show that in every execution, 
if $n_e$ enqueues and $n_d$ dequeues are invoked, the total number
of steps taken by all processes is $O(n_e\cdot T_e + n_d\cdot T_d)$.
}

>>>>

\rev{
We use a  standard model of an asynchronous, shared-memory system.
A collection of $p$ processes can communicate by accessing
shared memory locations using read, write and compare-and-swap (\op{CAS}) instructions.
A \op{CAS}(\var{X}, \var{old}, \var{new}) instruction updates a variable \var{X} to the value \var{new} and returns \tr\ if \var{X}'s value is \var{old}; otherwise the \op{CAS} leaves \var{X} unchanged and returns \fa.
We also assume processes can allocate a block of shared memory to represent an object.

We are concerned with implementing a first-in, first-out queue in such a system.
An implementation provides algorithms that a process can execute to perform
an enqueue or dequeue operation.
We treat individual memory instructions as atomic, so an execution
of such an implementation can be thought of as a sequence of steps by processes
where in each step a process performs at most one shared-memory instruction 
(and receives the response of that instruction).
The subsequence of steps performed by a process should follow that process's
algorithms for performing the process's desired sequence of enqueue and dequeue operations.
The order in which the steps of processes are interleaved is arbitrary, and the implementation must be correct for all
possible interleavings.

@@

The step complexity of an operation is the number of steps it performs.
The amortized step complexity is the total number of steps by all processes
divided by the number of operations they perform.
More formally, to show that the amortized step complexity is $O(T)$ per operation, we must show that in every execution, 
if $m$ enqueue and dequeue operations are invoked, the total number
of steps taken by all processes is $O(m\cdot T)$.
}
